\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-08-26 paper.qmd
Updated MNIST intro text

The MNIST dataset consists of $70000$ grayscale images of handwritten digits [@lecun2010]. @yingfan2021 used this dataset to evaluate how PaCMAP preserves local structures in \gD{}. Because of the large number of images, we selected the handwritten digit 1 which has a non-linear structure in \gD{}. There are $7877$ images. Also, applied PCA as a preprocessing step and selected first $10$ principal components. The objective is to evaluate whether PaCMAP preserves the non-linear structure in \gD{}.

@@

<!--PaCMAP param: n_components=2, n_neighbors=10, init=random, MN_ratio=0.9, FP_ratio=2.0-->
```{r}
#| fig-cap: "$2\\text{-}D$ layout from PaCMAP applied for the digit 1 of the MNIST dataset. The (hyper-)parameter settings, beyond the defaults, are $10$ nearest neighbors, ratio of the number of mid-near pairs to the number of neighbors equal to $0.5$, and the ratio of the number of further pairs to the number of neighbors is $2$. We use our *model-in-the-data-space* to assess whether this is an accurate representation of structure present in $10\\text{-}D$, or if it is misleading. The angle of the digit 1 varies along this structure. Images at the top-left of the $2\\text{-}D$ layout show the digit 1 angled more to the left, while images at the bottom-right show the digit 1 angled more to the right."
#| label: fig-pacmap-author
#| fig-pos: H
#| out-height: 22%

>>>>

The MNIST dataset consists of $70000$ grayscale images of handwritten digits [@lecun2010]. @yingfan2021 used this dataset to evaluate how PaCMAP preserves local structures in \gD{}. Because of the large number of images, the focus was specifically on the handwritten digit 1, which exhibits a non-linear structure in \gD{} (@fig-pacmap-author). There are $7877$ images of the digit 1 in the dataset. Additionally, PCA was applied as a preprocessing step and the first $10$ principal components were selected. The objective was to assess whether PaCMAP effectively preserves the non-linear structure in \gD{}.

@@

<!--PaCMAP param: n_components=2, n_neighbors=10, init=random, MN_ratio=0.9, FP_ratio=2.0-->
```{r}
#| fig-cap: "$2\\text{-}D$ layout from PaCMAP applied for the digit 1 of the MNIST dataset. The (hyper-)parameter settings, beyond the defaults, are $10$ nearest neighbors, ratio of the number of mid-near pairs to the number of neighbors equal to $0.5$, and the ratio of the number of further pairs to the number of neighbors is $2$. We use our *model-in-the-data-space* to assess whether this is an accurate representation of structure present in $10\\text{-}D$, or if it is misleading. The angle of the digit 1 images varies along this structure. Images at the top-left of the $2\\text{-}D$ layout show the digit 1 angled more to the left, while those at the bottom-right show the digit 1 angled more to the right."
#| label: fig-pacmap-author
#| fig-pos: H
#| out-height: 22%

### 2024-08-26 paper.tex
Updated MNIST intro text

The MNIST dataset consists of \(70000\) grayscale images of handwritten
digits \citep{lecun2010}. \citet{yingfan2021} used this dataset to
evaluate how PaCMAP preserves local structures in \gD{}. Because of the
large number of images, we selected the handwritten digit 1 which has a
non-linear structure in \gD{}. There are \(7877\) images. Also, applied
PCA as a preprocessing step and selected first \(10\) principal
components. The objective is to evaluate whether PaCMAP preserves the
non-linear structure in \gD{}.

@@

\caption{\label{fig-pacmap-author}\(2\text{-}D\) layout from PaCMAP
applied for the digit 1 of the MNIST dataset. The (hyper-)parameter
settings, beyond the defaults, are \(10\) nearest neighbors, ratio of
the number of mid-near pairs to the number of neighbors equal to
\(0.5\), and the ratio of the number of further pairs to the number of
neighbors is \(2\). We use our \emph{model-in-the-data-space} to assess
whether this is an accurate representation of structure present in
\(10\text{-}D\), or if it is misleading. The angle of the digit 1 varies
along this structure. Images at the top-left of the \(2\text{-}D\)
layout show the digit 1 angled more to the left, while images at the
bottom-right show the digit 1 angled more to the right.}

>>>>

The MNIST dataset consists of \(70000\) grayscale images of handwritten
digits \citep{lecun2010}. \citet{yingfan2021} used this dataset to
evaluate how PaCMAP preserves local structures in \gD{}. Because of the
large number of images, the focus was specifically on the handwritten
digit 1, which exhibits a non-linear structure in \gD{}
(Figure~\ref{fig-pacmap-author}). There are \(7877\) images of the digit
1 in the dataset. Additionally, PCA was applied as a preprocessing step
and the first \(10\) principal components were selected. The objective
was to assess whether PaCMAP effectively preserves the non-linear
structure in \gD{}.

@@

\caption{\label{fig-pacmap-author}\(2\text{-}D\) layout from PaCMAP
applied for the digit 1 of the MNIST dataset. The (hyper-)parameter
settings, beyond the defaults, are \(10\) nearest neighbors, ratio of
the number of mid-near pairs to the number of neighbors equal to
\(0.5\), and the ratio of the number of further pairs to the number of
neighbors is \(2\). We use our \emph{model-in-the-data-space} to assess
whether this is an accurate representation of structure present in
\(10\text{-}D\), or if it is misleading. The angle of the digit 1 images
varies along this structure. Images at the top-left of the
\(2\text{-}D\) layout show the digit 1 angled more to the left, while
those at the bottom-right show the digit 1 angled more to the right.}
